Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-19 aistats.tex
Add TODOs in appendix

These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
\citep[p.177]{nesterov2003introductory}
, but $\logpart$ is not (see App.~\ref{app:gaussian}).
We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.

@@

To get an upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focused on the symmetrized Bregman instead
\aligns{
	\label{eq:log_bound}
	\phi(z) \leq \phi(z) + \phi(\inv{z}) 
	&=  \half \paren{z -1  + \inv{z} -1} 
	= \frac{(z-1)^2}{z}
}
\aligns{
	\bregmanconj( \mu_*; \mu_n) 
	&\leq \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) \\ 
	=: \cS_{\conj} (\mu_*, \mu_n )  
	&= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 }
}

@@

\subsection{Bounds on the Expected MAP Natural Parameter}
Before proving a convergence rate for the MAP, we need to bound the expectation of its inverse, hence the following lemma.
We introduce the notation $(z)_+ = \max(0,z)$.
\begin{lemma}[Expected MAP natural parameter]
	Define $a=n_0 \frac{\mu_0}{\mu^*} $ a variable characterizing the importance of the prior relative to the true parameters.
	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
\begin{equation}
		\frac{n_0+n}{a+n}
		=\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\nat^*}} 
		\leq
		\frac{n_0 +n}{a+(n-2)_+}
		\label{eq:lemma_nat_bound}
	\end{equation}
\end{lemma}

@@

\section{PROOFS FOR GAUSSIAN}
\label{app:gaussian}

@@

\section{SELF-CONCORDANCE}
\label{app:self-concordant}

@@

@@

>>>>

These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
\citep[p.177]{nesterov2003introductory} % example 4.1.1.4
, but $\logpart$ is not (see App.~\ref{app:gaussian}).
We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.

@@

To get an upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we will focus on the symmetrized Bregman instead
\aligns{
	\label{eq:log_bound}
	\phi(z) \leq \phi(z) + \phi(\inv{z}) 
	&=  \half \paren{z -1  + \inv{z} -1} 
	= \frac{(z-1)^2}{z}
}
\aligns{
	\bregmanconj( \mu_*; \mu_n) 
	&\leq \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) \\ 
	=: \cS_{\conj} (\mu_*, \mu_n )  
	&= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 }
}

@@

\subsection{Bounds on the Expected MAP Natural Parameter}
Before proving a convergence rate for the MAP, we need to bound the expectation of its inverse, hence the following lemma.
We introduce the notation $(z)_+ = \max(0,z)$.
\begin{lemma}[Expected MAP natural parameter]
	Define $a=n_0 \frac{\mu_0}{\mu^*} $ a variable characterizing the importance of the prior relative to the true parameters.
	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
\begin{equation}
		\frac{n_0+n}{a+n}
		=\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\nat^*}} 
		\leq
		\frac{n_0 +n}{a+(n-2)_+} \; , \forall n\geq 0 \; .
		\label{eq:lemma_nat_bound}
	\end{equation}
\end{lemma}

@@

\subsection{Proof of MAP Bound}

\begin{theorem}[MAP Bound]
The expected symmetrized Bregman of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is upper bounded as
 \begin{align}
	& \expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
		\inv{2(n_0+1)}  +  b_1 \ \text{ if }\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{ if }\ n\geq 2
	\end{cases}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}

\begin{proof}
When $n=0$, the inequality is an equality. 
Wen $n>0$, Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus on the simple log-bound given by \eqref{eq:log_bound},
\begin{align}
	2\expect{\bregmanconj(\mu^*; \MAPm)} 
	&\leq \expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1 \; .
\end{align}
We readily get 
\begin{align}
	 \expect{\frac{\MAPm}{\mu^*}} - 1 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
\end{align}
where recall that $a=n_0\frac{\mu_0}{\mu^*}$. 
There remains the more problematic term with the expectation of the inverse mean parameter.
Fortunately, we derived the bound \eqref{eq:lemma_nat_bound} for this purpose.

\textbf{When $n\geq 2$}, we get 
\begin{align}
	\expect{\frac{\mu^*}{\MAPm}} - 1 
	\leq \frac{n_0 + n}{a +n - 2} -1
	= \frac{n_0 - a + 2}{a +n - 2}
	 = \frac{2}{a +n - 2} + \frac{n_0 - a}{a +n - 2}
\end{align}
so putting it all together we get 
\begin{align}
	\expect{\bregmanconj(\mu^*; \MAPm)} 
	&\leq \half \left(\frac{n_0 - a + 2}{a +n - 2}  + \frac{a - n_0}{n_0+n} \right )\\
	&= \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
	&= \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2}{2(a+n-2)(n_0 + n)} \\
	&= \frac{1}{a+n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
\end{align}

\textbf{When $n=1$} the bound~\eqref{eq:lemma_nat_bound}  on the expected natural parameter gives
\begin{align}
	\expect{\frac{\mu^*}{\MAPm}} - 1 
	\leq \frac{n_0 + 1}{a} -1
	= \frac{n_0 +1 - a}{a}
\end{align}
so putting it all together we get
\begin{align}
	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 +1 - a }{a} \\
	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{\frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
\end{align}
so we do recover the same bias variance structure as when $n\geq2$.
\end{proof}

\subsection{Complementary Plots}

\rlp{Show that the optimal mu0 is larger than mu* for small n0.}

\section{COMPLEMENTS ON GAUSSIANS}
\label{app:gaussian}

\rlp{prove that $A$ is not self -concordant, while, $A^*$ is thanks to Nesterov.}
Maybe plot both of them.
Also plot the mirror maps.
And plot gaussian trajectories.

@@

\rlp{simple stuff}

\section{SELF-CONCORDANCE}
\label{app:self-concordant}
\TODO{Damien : review self-concordance + prove proposition 1 + graph to explain $x<0.21$.}

\rlp{Add connection duality / self-concordant}

@@

\rlp{prove that the bias decreases like $1/n^2$ for gaussian variance and gaussian mean.}

\rlp{prove that the expectation of Hanzely is the symmetrized bias, e.g.  an upper bound on the bias.}

@@

\rlp{Introduce SMD, in primal and dual formulations.}
